---
title: DEG analysis of LINCS compounds in all cell lines to get LINCS compounds signature
author: "Author: Yuzhu Duan"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 4
    fig_caption: yes
    code_folding: show
    number_sections: true

fontsize: 14pt
bibliography: bibtex.bib
---

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('sample.Rmd', output_format=c('pdf_document', 'html_document'), clean=TRUE)"; Rscript -e "knitr::knit('sample.Rmd', tangle=TRUE)"

- Commit to github
git add -A :/ 
git commit -am "some edits"; git push -u origin master 

- To customize font size and other style features, add this line to output section in preamble:  
    css: style.css
-->

# Get LINCS signatures for all LINCS compounds in every LINCS cell lines.

## Get level 3 quantile nomalized gene expression data of lincs compounds treatment and control sample in all cell lines.

**Only 812/2512 lincs compounds treatment have normalized gene expression data in `GEO70138` according to compounds name consistency**

treatment as `cmp_cell`, control as `DMSO_cell`.  There are 2039 samples for `DMSO_A375` control. Here, randomly select 50 DMSO samples for control.

Get gene expression data of pertabation treatment samples
```{r ,eval=FALSE}
q("no") # closes R session on head node
srun --x11 --partition=girkelab --mem=20gb --cpus-per-task 8 --ntasks 1 --time 48:00:00 --pty bash -l
R
system("hostname")
library(rhdf5); library(tidyverse)
source("io_new.R"); source("utils.R")
inst_info <- read.delim("data/GSE70138_Broad_LINCS_inst_info_2017-03-06.txt",sep = "\t", header = T, row.names = 1) # 345976 12
gctx_rids <- read.gctx.ids("data/GSE70138_Broad_LINCS_Level3_INF_mlr12k_n345976x12328_2017-03-06.gctx")
gctx_colids <- read.gctx.ids("data/GSE70138_Broad_LINCS_Level3_INF_mlr12k_n345976x12328_2017-03-06.gctx", dimension = "col") # 345976

pert_info_total <- as_tibble(read.delim("data/lincs_pert_info_total.txt", header = T, sep = "\t", stringsAsFactors=FALSE))  # 2836 12
# [1] "canonical_smiles"  "description"       "inchi_key"         "inchi_string"      "molecular_formula" "pert_id"           "pert_iname"        "pert_type"        
# [9] "pert_vendor"       "pubchem_cid"       "structure_url"     "id"       
length(unique(pert_info_total$pert_iname)) # 2512
length(unique(pert_info_total$pert_id)) # 2836
sum(pert_info_total$pert_id %in% inst_info$pert_id) # 711
sum(unique(pert_info_total$pert_iname) %in% inst_info$pert_iname) # 812
pert <- unique(pert_info_total$pert_iname)[(unique(pert_info_total$pert_iname) %in% inst_info$pert_iname)] # 812

pert_l3_inst_info <- filter(inst_info, inst_info$pert_iname %in% pert_info_total$pert_iname) # 158507 X 12
cell_type <- as.character(unique(pert_l3_inst_info$cell_id)) # 30
pert_l3_expr <- parse.gctx("data/GSE70138_Broad_LINCS_Level3_INF_mlr12k_n345976x12328_2017-03-06.gctx", cid = pert_l3_inst_info$inst_id)
pert_l3_expr <- pert_l3_expr@mat # 12328 X 158507
pert_l3_expr <- as.data.frame(pert_l3_expr)
write.table(pert_l3_expr, file="data/pert_l3_expr.xls", quote=FALSE, sep="\t", col.names = NA)
pert_l3_expr <- read.delim("data/pert_l3_expr.xls", header = T, row.names = 1, sep="\t") 
```

Get gene expression data of DMSO control samples.
```{r dmso, eval=FALSE}
dmso_inst <- filter(inst_info, pert_iname == "DMSO") # 19258 X 12
dmso_l3_expr <- parse.gctx("data/GSE70138_Broad_LINCS_Level3_INF_mlr12k_n345976x12328_2017-03-06.gctx", cid = dmso_inst$inst_id)
dmso_l3_expr <- as.data.frame(dmso_l3_expr@mat) # 12328 X 19258
write.table(dmso_l3_expr, "data/dmso_l3_expr.xls", col.names = NA, quote = FALSE, sep = "\t")
pert_dmso_l3_expr <- cbind(pert_l3_expr, dmso_l3_expr) # 12328 X 177765
write.table(pert_dmso_l3_expr, file="data/pert_dmso_l3_expr.xls", quote=FALSE, sep="\t", col.names = NA)
```

Log2 the L1000 normalized gene expression value of pertabation treatment sample and dmso control sample, set -Inf to -1000000, store the result as `pert_dmso_l3_expr_log2_noinf.xls`. 
```{r log_noinf, eval=FALSE}
library(data.table)
pert_dmso_l3_expr <- fread("data/pert_dmso_l3_expr.xls", header=TRUE, sep="\t")
pert_dmso_l3_expr <- as.data.frame(pert_dmso_l3_expr)
rownames(pert_dmso_l3_expr) <- as.character(pert_dmso_l3_expr$"V1")
pert_dmso_l3_expr <- pert_dmso_l3_expr[,2:dim(pert_dmso_l3_expr)[2]]
df <- log2(pert_dmso_l3_expr)
df_mat <- as.matrix(df)
df_mat[df_mat == -Inf] <- -1000000
df_noinf <- as.data.frame(df_mat)
saveRDS(df, "data/pert_dmso_l3_expr_log2.rds")
write.table(df, file="data/pert_dmso_l3_expr_log2.xls", quote=FALSE, sep="\t", col.names = NA)
write.table(df_noinf, file="data/pert_dmso_l3_expr_log2_noinf.xls", quote=FALSE, sep="\t", col.names = NA)
```

## Creat comp_list. In each cell type, randomly select 50 DMSO samples as control.
```{r comp_list,eval=FALSE}
pert_dmso_inst_info$cmp_cell <- paste(pert_dmso_inst_info$pert_iname, pert_dmso_inst_info$cell_id, sep="_")
pert_l3_inst_info$cmp_cell <- paste(pert_l3_inst_info$pert_iname, pert_l3_inst_info$cell_id, sep="_")

comp_list <- list()
for(item in unique(pert_l3_inst_info$cmp_cell)){
  comp_list[[item]] <- list()
  comp_list[[item]]$t <- filter(pert_dmso_inst_info, cmp_cell == item)$inst_id
  cell <- unlist(strsplit(item, "_"))[2]
  comp_list[[item]]$c <- filter(pert_dmso_inst_info, cmp_cell == paste0("DMSO_", cell))$inst_id
} # 7327
saveRDS(comp_list, "data/comp_list.rds")
comp_list <- readRDS("data/comp_list.rds")
```

## DEG analysis with `limma` to get signatures

```{r deg_limma,eval=FALSE}
comp_list <- readRDS("data/comp_list.rds") # 7327
comp_list_c50 <- comp_list
for(name in names(comp_list_c50)){
  if(length(comp_list_c50[[name]]$c)>=50){
    comp_list_c50[[name]]$c <- sample(comp_list_c50[[name]]$c, 50)
  }
}
df_noinf <- read.delim("data/pert_dmso_l3_expr_log2_noinf.xls", sep="\t", header=T, row.names=1, check.names=FALSE)
library(longevityTools)
degList <- runLimma(df=df_noinf, comp_list=comp_list_c50, fdr=0.10, foldchange=1, verbose=TRUE, affyid=NULL)

write.table(degList$DEG, file="./data/degMA.xls", quote=FALSE, sep="\t", col.names = NA) 
saveRDS(degList$DEG, "./data/degMA.rds") # saves binary matrix 
saveRDS(degList, "./data/degList.rds") # saves entire degList
```

Number of DEGs across drug treatments
```{r deg_plot,eval=TRUE}
degMAgene <- readRDS("./data/degMA.rds")
y <- as.numeric(colSums(degMAgene))
interval <- table(cut(y, right=FALSE, dig.lab=5,  breaks=c(0, 5, 10, 50, 100, 200, 500, 1000, 10000)))
df <- data.frame(interval); colnames(df) <- c("Bins", "Counts")
library(ggplot2)
ggplot(df, aes(Bins, Counts)) + 
       geom_bar(position="dodge", stat="identity", fill="cornflowerblue") + 
       ggtitle("DEG numbers by bins")
```


